The Way Forward in Parallelizing Dynamic Languages

### 2014-04-29 talk/icooolps2014/position-paper.tex
xxx: should mention that the interpreter is typically very large and maintained by open-source communities
more text

\section{Introduction}
In a world where computers get more and more cores and single-thread
performance increases less and less every year, many dynamic languages
have a problem. While there is certainly a lot of popularity around
languages like Python and Ruby, their ability to make use of multiple
cores is somewhat limited. For ease of implementation they chose to
use a single, global interpreter lock (GIL) to synchronize the
execution of code in multiple threads. While this is a
straight-forward way to eliminate synchronization issues in the
interpreter, it prevents parallel execution. Code executed in multiple
threads will be serialized over this GIL so only one thread executes
at a time.

@@

These requirements are not easy to meet. We argue that STM is the
overall winner. While it has a big performance problem currently, it
gets more points in all the other categories. We think that it is the
only solution that also enables a better synchronization mechanism to
the application in the form of atomic blocks.

@@

\section{Discussion}
\subsection{Why is there a GIL?}
The GIL is a very simple synchronization mechanism for supporting
multi-threading in the interpreter. The basic guarantee is that the
GIL may only be released in-between bytecode instructions. The
interpreter can thus rely on complete isolation and atomicity of these
instructions. As a consequence, applications can rely on certain
operations to be atomic. While this is probably not a good idea,
it is used in practice. A solution replacing the GIL should therefore
uphold these guarantees, while preferably also be as easily
implementable as a GIL for the interpreter.

@@

\paragraph{GIL}

- nice semantics\\
- easy support of atomic blocks\\
- no parallelism

\paragraph{fine-grained locking}

@@

\paragraph{multiprocessing / no-sharing models}

@@

\paragraph{HTM}

@@

\section{Potential Approach}
possible solution:\\
- use virtual memory paging to somehow lower the STM overhead\\
- tight integration with GC and jit?

>>>>

\section{Introduction}
In a world where computers get more and more cores and single-thread
performance increases less and less every year, many dynamic languages
have a problem. While there is certainly a lot of popularity around
languages like Python and Ruby, their ability to make use of multiple
cores is somewhat limited. For ease of implementation they chose to
use a single, global interpreter lock (GIL) to synchronize the
execution of code in multiple threads. While this is a
straight-forward way to eliminate synchronization issues in the
interpreter, it prevents parallel execution. Code executed in multiple
threads will be serialized over this GIL so that only one thread can
execute at a time.

@@

These requirements are not easy to meet. We argue that STM is the
overall winner. While it has a big performance problem currently, it
gets more points in all the other categories. We think that it is the
only solution that also provides a better synchronization mechanism to
the application in the form of atomic blocks.

@@

\section{Discussion}

\subsection{Why is there a GIL?}
The GIL is a very simple synchronization mechanism for supporting
multi-threading in the interpreter. The basic guarantee is that the
GIL may only be released in-between bytecode instructions. The
interpreter can thus rely on complete isolation and atomicity of these
instructions. As a consequence, applications can rely on certain
operations to be atomic. While this is probably not a good idea,
it is used in practice. A solution replacing the GIL should therefore
uphold these guarantees, while preferably also be as easily
implementable as a GIL for the interpreter.

@@

The list of criterias for evaluating potential solutions is as follows:
\begin{itemize}
\item Performance
\item Changes required to existing applications
\item Better synchronization options for applications (e.g. atomic blocks)
\item Ease of implementation (interpreter-level)
\item Integration with external libraries
\end{itemize}

\subsubsection*{fine-grained locking}

@@

\subsubsection*{Shared-nothing / multiple processes}

@@

\subsubsection*{Transactional Memory}
\paragraph{HTM}

@@

\section{The Way Forward}
possible solution:\\
- use virtual memory paging to somehow lower the STM overhead\\
- tight integration with GC and jit?
